---
title: "Assignment 4 demo"
author: "Roger Day"
date: "February 24, 2016"
output: pdf_document
---

Demonstration of Central Limit Theorem successes and failures
===

IMPORTANT NOTE:  see also CLTdemo.Rmd to update this exercise.

# Purpose
For each of several distributional families, we will generate many samples at various sample sizes, in order to characterize how the distribution of the sample mean changes with sample size.

# Families and parameters
```{r}
familyList = list(
    list(family="rbinom", extraArgs=list(size=1, prob=0.2)),
    list(family="rpois", extraArgs=list(lambda = 1)),
    list(family="rnorm", extraArgs=list(mean = 0, sd = 1)),
    list(family="rexp", extraArgs=list()),
    list(family="rcauchy", extraArgs=list())
)
```

# Sample sizes to test, and number of replicatios
```{r}
sampleSizeVector = c(1, 10, 100)
nReps = 1e4
```

The following function tests a single sample size, returning a vector of sample means.
```{r}
CLTtestOneSize = function(distribution, sampleSize, nReps) {
    numbers = do.call(
      what = distribution$family, 
      args = c(n = sampleSize*nReps, distribution$extraArgs)
    )
    # Each column is a sample
    simulations = matrix(numbers, nrow=sampleSize)
    # return(apply(simulations, 2, mean))  # one way
    return(colMeans(simulations))  # simpler
}
```

This function assembles the samples of means across the sample sizes, and plots all on a single plot. The "..." represents any unmatched args. They are passed into the call to plot().
```{r}
CLTdensityPlot = function(distribution, theXrange, ...) {
  densitiesOfMeans = lapply(
    sampleSizeVector, 
    function(sampleSize) {
      simMeans = CLTtestOneSize(
        distribution, sampleSize, nReps)
      return(density(simMeans, kernel="triangular", bw=0.02))
    }
  )
  names(densitiesOfMeans) = sampleSizeVector
if(missing(theXrange))
  theXrange = range(c(sapply(
    densitiesOfMeans, function(density) range(density$x)) ))
  theYrange = range(c(sapply(densitiesOfMeans, 
                      function(density) range(density$y)) ))
  colors = 1 + 1:length(sampleSizeVector)
  chars = 1 + 1:length(sampleSizeVector)
  for(i in 1:length(sampleSizeVector)) {
    if(i==1)
      plot(densitiesOfMeans[[i]],  type="b"
           , col=colors[1], pch=chars[1]
           , xlim=theXrange
           , ylim=theYrange
           , xlab="X=x"
           , ylab="Density"
           , main="" 
           , ...
      )
    else points(densitiesOfMeans[[i]], type="b",
                col=colors[i], pch=chars[i])
  }
  title(main=paste(distribution$family, "\n",
      paste(names(distribution$extraArgs), 
                          distribution$extraArgs,
            sep = "=", collapse = ", ")))

legend(x="topright",legend=sampleSizeVector,
       col=colors, pch=chars)
}
```
We'll jump to a new page now.
For a pdf document, it's done with a **TeX** command,
> \\newpage 

and with an html document, it's done with an **HTML** command, 
> <br>

\newpage

#The `r familyList[[1]]$family` family
```{r}
CLTdensityPlot(familyList[[1]])
```



For the `r familyList[[1]]$family` family,
we see that the distribution of the mean converges to a normal distribution.

Rerunning this chunk gives similar-looking results each time.
\newpage

#  For the `r familyList[[5]]$family` family:
```{r}
CLTdensityPlot(familyList[[5]], theXrange=c(-500, 500))
```

For the `r familyList[[5]]$family` family,
we see that the distribution of the mean does NOT converge to a normal distribution, because the Cauchy distribution does not have a defined mean (and variance).

Repeating this chunk will give very different results each time, because most of the action is far out on the tails.

