Confidence intervals for a binomial.
===

Normal approximation
```{r}
0.8+c(-1,1)*sqrt(.8*.2/9)*qnorm(0.975)
```

Compare to exact.  Here's a function that uses the duality between tests and confidence intervals. It handles two-sided, upper, and lower intervals.
```{r}
binom.confint.new = 
function(k, n, alpha=0.05, range=0:1, 
		side=c("two", "upper", "lower")) {
	sideLetter = toupper(substring(side[1], 1, 1))
	if(sideLetter=="T") {
	  alphaUpper = alphaLower = alpha/2
	} else if(sideLetter=="U") {
	  alphaUpper = alpha; alphaLower = 0
	} else if(sideLetter=="L") {
	  alphaLower = alpha; alphaUpper = 0
	}
	else stop("bad value for \"side\" ")  	
	huntForBoundaryUpper = function(p) 
	  alphaUpper - (1 - pbinom((k-1),n,p))
	lb = try(uniroot(huntForBoundaryUpper, range)$root)
	if(class(lb) == "try-error") lb = 0
	huntForBoundaryLower = function(p) 
  			alphaLower - pbinom(k,n,p)
  	ub = try(uniroot(huntForBoundaryLower, range)$root)
  	if(class(ub) == "try-error") ub = 1
	return(c(lb=lb, ub=ub))
}
```
Now let's compare the three confidence intervals.
```{r}
plot(c(0, 1), 0:1, pch="", xlab="P", ylab="", 
     xaxs="i", yaxt="n")
abline(v=8/10, col="red")
lines(binom.confint.new (k=8, n=10), c(1/4,1/4),
      pch="2", type="b")
text(0.8, 1/4, adj=c(1,1), col="darkgreen", "two-sided ")
lines(binom.confint.new (k=8, n=10, side="u"), c(2/4,2/4),
      pch="U", type="b", xpd=NA)
text(0.8, 2/4, adj=c(1,1), col="darkgreen", "one-sided (upper) ")
lines(binom.confint.new (k=8, n=10, side="l"), c(3/4,3/4),
      pch="L", type="b", xpd=NA)
text(0.8, 3/4, adj=c(1,1), col="darkgreen", "one-sided (lower) ")

```


NOTES
---
* The (badly named) parameter xpd governs clipping. We don't want the letters to be clipped.
* The xaxs="i" parameter forces the x axis to match the xlim exactly.
* The yaxt="n" means no ticks or labels on the y axis.