title: "Confidence intervals for a binomial"
author: "Roger Day"
date: "2/15/2018"
output: html_document
runtime: shiny

===

Normal approximation
```{r}
0.8+c(-1,1)*sqrt(.8*.2/9)*qnorm(0.975)
```

Compare to exact.  Here's a function that uses the duality between tests and confidence intervals. It handles two-sided, upper, and lower intervals.
```{r}
binom.confint.new = 
function(k, n, alpha=0.05, range=0:1, 
		side=c("two", "upper", "lower"), traces=TRUE) {
	sideLetter = toupper(substring(side[1], 1, 1))
	if(sideLetter=="T") {
	  alphaUpper = alphaLower = alpha/2
	} else if(sideLetter=="U") {
	  alphaUpper = alpha; alphaLower = 0
	} else if(sideLetter=="L") {
	  alphaLower = alpha; alphaUpper = 0
	}
	else stop("bad value for \"side\" ")  	
	saved_p_u <<- numeric(0)
	huntForBoundaryUpper = function(p) {
	  saved_p_u <<- c(saved_p_u, p)
	  alphaUpper - (1 - pbinom((k-1),n, p))
	}
	lb  = try({
	  rootHuntUpper <<- uniroot(huntForBoundaryUpper, range);
	  rootHuntUpper$root
	})
	
	if(class(lb) == "try-error") lb = 0
	saved_p_l <<- numeric(0)
	
	huntForBoundaryLower = function(p) {
	  saved_p_l <<- c(saved_p_l, p)
	  alphaLower - pbinom(k,n,p)
	}
	ub = try({
	  rootHuntLower <<- uniroot(huntForBoundaryLower, range);
	  rootHuntLower$root
	})
	if(class(ub) == "try-error") ub = 1
	assign('saved_p_u', saved_p_u)
	return(c(lb=lb, ub=ub))
}
```


Now let's compare the three confidence intervals.
```{r}
plot(c(0, 1), 0:1, pch="", xlab="P", ylab="", 
     xaxs="i", yaxt="n")
abline(v=8/10, col="red")
lines(binom.confint.new (k=8, n=10, side="u"), c(2/4,2/4),
      pch="U", type="b", xpd=NA)
text(0.8, 2/4, adj=c(1,1), col="darkgreen", "one-sided (upper) ")
lines(binom.confint.new (k=8, n=10, side="l"), c(3/4,3/4),
      pch="L", type="b", xpd=NA)
text(0.8, 3/4, adj=c(1,1), col="darkgreen", "one-sided (lower) ")
lines(binom.confint.new (k=8, n=10), c(1/4,1/4),
      pch="2", type="b")
text(0.8, 1/4, adj=c(1,1), col="darkgreen", "two-sided ")

```


NOTES
---
* The (badly named) parameter xpd governs clipping. We don't want the letters to be clipped.
* The xaxs="i" parameter forces the x axis to match the xlim exactly.
* The yaxt="n" means no ticks or labels on the y axis.

### A diagram showing the tail probabilities.

## First for the lower bound (upper tail):
```{r}
for(p_u in saved_p_u) {
    plot(0:10, dbinom(0:10, 10, p_u)
        ,type='h' 
       , col=c('black','red')[1 + (0:10 >= 8)]
       )
  abline(v=p_u*10, col='green', lwd=4)
  title(paste('P = ', p_u, '\n',
              'upperTail=', 1-pbinom(8-1,10,p_u)))
}

```

## Then for the upper bound (lower tail):
```{r}
for(p_l in saved_p_l) {
    plot(0:10, dbinom(0:10, 10, p_l)
        ,type='h' 
       , col=c('black','red')[1 + (0:10 <= 8)]
       )
  abline(v=p_l*10, col='green', lwd=4)
  title(paste('P = ', p_l, '\n',
              'lowerTail=', pbinom(8,10,p_l)))
}

```

And as a reminder, the confidence interval (two-sided) is:
```{r}
binom.confint.new (k=8, n=10)
```

