---
title: " chisq-voting-table.Rmd"
author: "roger"
date: "3/20/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

####   Example 9.3.1  from DeGroot-Schervish.

To set up the data, one can select the contents of the "Observations" table in "N15-categorical data.docx", and paste it into a string, here called `voterDataString`. Then write string to a temporary file, and read it into a data frame.
```{r}
voterDataString <- 
  "	Prefers.Mr.Jones	Prefers.Ms.Smith	undecided	TOTAL
Eng and science	24	23	12	59
Humanities, soc sciences	24	14	10	48
Fine arts	17	8	13	38
Industr, public admin	27	19	9	55
TOTAL	92	64	44	200"
temp <- tempfile()
cat(voterDataString, file=temp)
voterData <- read.delim(temp, sep = '\t')
str(voterData)
```

Now we have to use the first column as the rownames, remove the first column, and remove the row and column labeled TOTAL.

```{r}
rownames(voterData) <- voterData[ , 1]
voterData <- voterData[ , -1]  ## Remove first column (used for row names)
# It still has the TOTAL rows and columns. We must remove them.
voterData <- voterData[ , -match("TOTAL", colnames(voterData))]  ## Remove first column (used for row names)
voterData <- voterData[ -match("TOTAL", rownames(voterData)), ]  ## Remove first column (used for row names)
voterData
```


Now we want to fit models using the function `loglin()`.
But voterData is a data frame.  
To use loglin() for log-linear modeling, it must be a matrix.
```{r}
voterData <- as.matrix(voterData)
voterData   #### OK
longDimNames <- dimnames(voterData)
dimnames(voterData) <- list(c("E","H","F","I") , 
                            c("Jones", "Smith","un"))
str(voterData)
```

Later we'll change one number dramatically, and re-knit this document, to see what happens. We use the `eval` arg in the chunk clause to control this. 

```{r eval=FALSE}
voterData[1,2] <- 230  
```

Meanwhile, let's try some log-linear modeling. The `margin` argument says what "faces" are participating in the model. Here, `list(1,2)` means that the model will have row effects (1) and column effects (2), but not the interaction.
```{r}
voterData.fitted <- loglin(table = voterData, margin = list(1,2), fit=TRUE)
voterData - voterData.fitted$fit   ##  O - E
(voterData.fitted$fit - voterData)^2/voterData.fitted$fit
(voterData.fitted$fit - voterData)/sqrt(voterData.fitted$fit)

Q <- sum((voterData.fitted$fit - voterData)^2/voterData.fitted$fit)
c(Q=Q, pearson=voterData.fitted$pearson, lrt=voterData.fitted$lrt)

df.for.Q <- (nrow(voterData)-1) * (ncol(voterData)-1)
c(voterData.fitted$df, df.for.Q)
cat("P = ", 1 - pchisq(Q, df.for.Q), "\n")
```
Interaction terms can be specified like this (to get the saturated model):
```{r}
loglin(table = voterData, margin = list(1, 2, c(1,2)), fit=TRUE)
```

Very handy for modeling higher order data.

### Using loglm() ###

Here is another approach, using the function `loglm()`, which uses the "formula" syntax to specify the model.  The first call gives us the saturated model.
```{r}
require("MASS")   ### "MASS" is a library in support of the book "Modern Applied Statistics with S-Plus", Venables and Ripley.
voterData.loglm.out <- loglm( ~ 1 + 2 + 1:2, voterData)
print(voterData.loglm.out)
voterData.loglm.out <- loglm( ~ 1 + 2, voterData)
print(voterData.loglm.out)
```

Ordinarily, in a formula object (created by the "~" character), the right side indicates predictors, and the left side is the predicted variable or "label".

For loglinear models, however, there is no left side, because the counts themselves are the things to predict.

###  Reconstructing the likelihood ratio statistic: ###
Depending on your computer, the likelihood might overflow!
So better to do the loglikelihood directly.

```{r}
maxLik1 <- prod(voterData^voterData)   # SATURATED
log(maxLik1)  #### It COULD overflow!
logmaxLik1 <- sum(voterData*log(voterData))  ### Better
logmaxLik0 <- sum(voterData*log(voterData.fitted$fit))
c(2*(logmaxLik1 - logmaxLik0),
  voterData.fitted$lrt )
```
 They agree, of course.

  What if you want to test whether the departments are different?  

  If not, then the null hypothesis is that all the row probabily vectors are IDENTICAL, not just proportional.

  So the d.f. is 8 = (RC-1) - (R-1) which equals R(C-1).

```{r}
voterData.loglin.out = loglin( voterData, margin = 1)
print(voterData.loglin.out)
1 - pchisq(voterData.loglin.out$pearson, 8)
```

Notice that functions like loglin, loglm, and regression functions in R do not automatically produce P values. I think they want to encourage you to know what you're doing!


