---
title: "eggs and the piercer"
output: html_notebook
---

#8.4 in Dalgaard

Reading the data.
```{r}
eggsTable = data.frame(
  size=c('A','B','A','B'), piercer=c('N','N','Y','Y'),
  total=c(54,200,60,70), 
  broken=c(4,15,4,1), 
  cracked=c(8,28,9,7))
eggsTable$neither = eggsTable$total - eggsTable$broken - eggsTable$cracked
eggsDF = expand.grid(size=c('A','B'),piercer=c('N','Y'),
                    result=c('broken','cracked','neither'))
eggsDF$n = c(eggsTable$broken, eggsTable$cracked, eggsTable$neither)
eggsDF
```
```{r}
eggsArray = array(eggsDF$n, 
                  dim=c(2,2,3), 
                  dimnames=list(
                    size=c('A','B'),piercer=c('N','Y'),
                    result=c('broken','cracked','neither')))
eggsArray
```

First, let's look at the design; is it balanced with respect to size?

```{r}
sizeByPiercer = apply(eggsArray, 1:2, sum)
testDesign = chisq.test(sizeByPiercer)
cat("P = ", testDesign$p.value,'\n')
testDesign$expected
testDesign$observed
testDesign$stdres
```

Highly imbalanced design.

Is "size" a "risk factor"?
```{r}
resultBySize = apply(eggsArray, c(1,3), sum)
testresultBySize = chisq.test( resultBySize)
cat("P = ", testresultBySize$p.value,'\n')
testresultBySize$expected
testresultBySize$observed
testresultBySize$stdres
```

Does the piercer affect broken-or-cracked?
This test ignores size. That's probably OK, given the last result. 

```{r}
resultByPiercer = apply(eggsArray, 2:3, sum)
testResultByPiercer = chisq.test( resultByPiercer)
cat("P = ", testResultByPiercer$p.value,'\n')
testResultByPiercer$expected
testResultByPiercer$observed
testResultByPiercer$stdres
```

Let's look for a "Simpson's Paradox".
We will do the analysis separately for A and B.


```{r}
for(size in c('A','B')) {
  resultByPiercerGivenSize = eggsArray[ size, , ]
  testresultByPiercerGivenSize = chisq.test( resultByPiercerGivenSize)
  cat("For size = ", size, ":\n")
  cat("P = ", testresultByPiercerGivenSize$p.value,'\n')
  cat("Fisher P = ", fisher.test(resultByPiercerGivenSize)$p.value,'\n')
  print(testresultByPiercerGivenSize$expected)
  print(testresultByPiercerGivenSize$observed)
  print(testresultByPiercerGivenSize$stdres)
}
```

## linear models adjusting for size

```{r}
glm(data=eggsDF, (result=="neither") ~ piercer,
    family=poisson, weight=n)
summary(.Last.value)
glm(data=eggsDF, (result=="neither") ~ size + piercer,
    family=poisson, weight=n)
summary(.Last.value)

```

## does the piercer affect broken | broken-or-cracked?
```{r}
glm(data=eggsDF, subset = (result!="neither"), 
    formula=(result=="broken") ~ piercer,
    family=poisson, weight=n)
summary(.Last.value)
eggsDFNotNeither = eggsDF[eggsDF$result!="neither", ]
glm(data=eggsDFNotNeither, 
    formula=(result=="broken") ~ size + piercer,
    family=poisson, weight=n)
summary(.Last.value)
```

By eyeball, it seems that possibly the "1" is low: 
the pierce might decrease broken eggs among B (big) eggs only.

To test this, we need interaction term. 

```{r}
glm(data=eggsDFNotNeither, 
    (result=="broken") ~ size * piercer,
    family=poisson, weight=n)
summary(.Last.value) 
```

Nope. nothing.

### The ANOVA approach, using MASS::loglm
```{r}
require(MASS)  ### Access to loglm()
eggsLogLm = loglm(formula = (result) ~  size + piercer,
                  data=eggsDF,
                  subset=(result!='neither'))
deviance(eggsLogLm)
coef(eggsLogLm)
fitted(eggsLogLm)
eggsAOV_H0 = glm(formula = (result) ~  size, 
                 family = 'binomial',
                 data=eggsDF[(eggsDF$result!="neither"), ], 
                 weights= n)
eggsAOV_HA = glm(formula = result ~  piercer + size, 
                 family = 'binomial',
                 data=eggsDF[(eggsDF$result!="neither"), ], 
                 weights= n)
anova(eggsAOV_H0, eggsAOV_HA)
deviance(eggsAOV_HA)
  
eggsAOV_HA = loglm(formula = result ~ piercer + size,
                   data=eggsDF, subset=result!="neither",
                   weights= n)
summary(eggsAOV_HA)
anova(eggsAOV_HA)
formulas = list(
  result ~ size,
  result ~ size + piercer,
  result ~ size * piercer)
glmresults = lapply(formulas, function(form)
  glm(formula=form,
       weights= nn, family = 'binomial',
       data=eggsDF, 
       subset=(eggsDF$result!="neither") 
       )
)
### having trouble just passing the args thru '...'.
sapply(glmresults, '[[', "deviance")
glmresultswithnames = glmresults
names(glmresultswithnames) = sapply(formulas, capture.output)
sapply(glmresultswithnames, '[[', "deviance")
# anova(glmresults)  fails; does not take a list,
# however this works:
do.call(what = anova, args = glmresults )

### Stepwise using AIC decides to drop everything!
step(glmresults[[3]])
```
