---
title: "The F distribution"
author: "roger"
date: "2/14/2017"
output: html_document
---
 The F distribution.  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
'%&%' = function(a,b) paste0(as.character(a),as.character(b))
```

See notes on testing variances.

We suppose six observations,  with sum of the squared deviations from the mean equalling 30,

and another 21 observations, with sum of squared deviations = 40.  

The ratio of the two variance estimates is (30/(6-1)) / (40/(21-1)) = `r (30/(6-1)) / (40/(21-1))`. 

The ratio of their square roots is sqrt(3) = `r sqrt(3)`.


### Testing

Could the 2 samples come from distributions with the same variance or not?  We use the F distribution.

```{r}
curve(df(x,5,20), xlim=c(0,8), yaxs="i")
curve(df(x,5,20), xlim=c(1/3,3), yaxs="i", add=T, type="h", col="green")
curve(df(x,5,20), xlim=c(0,1/3), yaxs="i", add=T, type="h", col="red")
curve(df(x,5,20), xlim=c(3,8), yaxs="i", add=T, type="h", col="red")

lines(c(1/3,1/3), c(0,df(1/3,5,20)), col="red")
lines(c(3,3), c(0,df(3,5,20)), col="red")

text(3, 0.4, "  Tails: 
     pf(1/3,5,20) + 1 - pf(3,5,20)
       = " %&% round(digits=3, pf(1/3,5,20) + 1 - pf(3,5,20)), 
     col="red")
```
The "tails of surprise" are:
```{r}
cat("lower tail: ", pf(1/3,5,20) , "\n")
cat("upper tail: ", 1 - pf(3,5,20) , "\n")
cat("both tails: ", pf(1/3,5,20) + 1 - pf(3,5,20) , "\n")
```
So one kind of two-sided  P-value is `r pf(1/3,5,20) + 1 - pf(3,5,20)`.


If we use density as our test statistic, the lower tail boundary is
```{r}
lowerboundary = uniroot(function(x) df(x,5,20)-df(3,5,20), 0:1)$root 
cat("lowerboundary = ", lowerboundary, "\n")
cat("lower tail: ", pf(lowerboundary,5,20) , "\n")
cat("P = ", 1 - pf(3,5,20) + pf(lowerboundary,5,20), "\n")
```
which is far smaller!

### Confidence intervals

A  confidence interval is obtained by requiring equal probability on each side.
Upper and lower 5% quantiles are:
```{r}
lowerbound = 3/qf(0.95, 5, 20) 
upperbound = 3/qf(0.05, 5, 20) 
```
Thus a confidence interval for the ratio of the variances is
```{r}
(c(lowerbound, upperbound))
```
The moment estimator is 3\ *(20-2)/20 = `r 3*(20-2)/20`.

Here we see the densities of the F for degrees of freedom (5,20) and (20,5). The vertical lines are at one and at the distribution mean.

We use `par(mfrow=c(1,2))` to put the plots side by side.
```{r}
par.saved = par(mfrow=c(1,2))
dfF = c(5,20)
for( dfF in list(c(5,20), c(20,5))) {
curve(from=0, to=2,  df(x, dfF[1], dfF[2]), type="l", main=paste("df = ", dfF[1], ",", dfF[2],
                                                                 "\nmean=", dfF[2], "/", (dfF[2]-2)), ylab="F density"); abline(v=c(1,dfF[2]/(dfF[2]-2)));
}
par(par.saved) ### reset.
```
