Demonstration of the likelihood ratio test
----

Setting: two independent binomial observations, X1 and X2.

The question: do they come from the same probability?

```{r fig.width=4, fig.height=4}
#par(fin=c(3,3))
n1 = 30 ; n2 = 40
p1 = 0.4; p2 = 0.7
X1 = 12; X2 = 28
#X1 = rbinom(1, n1, p1); X2 = rbinom(1, n2, p2)
symbols(1/2, 1/2, squares=0, add=FALSE, inches=F,
     xlab=expression(hat(p)[1]), xlim=0:1,
     ylab=expression(hat(p)[2]), ylim=0:1)
abline(a=0,b=1, lty=2, col="red", lwd=3)
points(c(X1/n1, (X1+X2)/(n1+n2)),
     c(X2/n2, (X1+X2)/(n1+n2)), cex=2,
     col=c("green", "red"), pch=16) #solid circle
points(c(X1/n1, (X1+X2)/(n1+n2)),
     c(X2/n2, (X1+X2)/(n1+n2)), cex=2) #open circle
text((X1+X2)/(n1+n2), (X1+X2)/(n1+n2),
     expression(group("(", list(hat(p),hat(p)), ")")),
     pos=4,
     col="red")
### pos=4  means   to the right side
text(X1/n1, X2/n2, 
     expression(group("(", list(hat(p)[1],hat(p)[2]), ")")), 
     pos=2,
     col="darkgreen")
```
 
Here is a function that does the LRT:
```{r}
lrt_2binomials = function(X1, X2, n1, n2) {
  p1hat = X1/n1; p2hat = X2/n2
  phat = (X1+X2)/(n1+n2)
  
  numerator = dbinom(X1, n1, p1hat) * dbinom(X2, n2, p2hat)
  denominator = dbinom(X1, n1, phat) * dbinom(X2, n2, phat)
  
  W = numerator/denominator
  
  2*log(W)  # twoLogLLR
}
```
For one replication,  the P value is 
   1 - pchisq(twoLogLLR, df=1) = 
     `r twoLogLLR=lrt_2binomials(X1, X2, n1, n2); 1 - pchisq(twoLogLLR, df=1) `

Now let's see what the power looks like:
We will repeat generating artificial data, and see how often the test rejects.
```{r}
lrt_2binomials_pvalues = function(p1, p2, n1, n2, N=10000) {
 sapply(1:N, function(ignoreMe) {
  X1 = rbinom(1, n1, p1)
  X2 = rbinom(1, n2, p2)
  1 - pchisq(lrt_2binomials(X1, X2, n1, n2), 1)
  }
 ) 
}
```
First, we'll check the type I error (power at H0)
```{r}
pvalues = lrt_2binomials_pvalues(0.3, 0.3, 30, 40)

```
For nominal alpha = 1%, the actual type I error is 
`r mean(pvalues < 0.01)`.

For nominal alpha = 5%, the actual type I error is `r mean(pvalues < 0.05)`.

Not too bad.

Now, let's check the power at an alternative, HA:
```{r}
pvalues = lrt_2binomials_pvalues(0.3, 0.6, 30, 40)
```
For nominal alpha = 1%, the power is `r mean(pvalues < 0.01)`.

For nominal alpha = 5%, the power is `r mean(pvalues < 0.05)`.

That's not terrific power, for detecting a doubling in the probability.
