---
title: "mcnemar exercise"
author: "roger"
date: "April 4, 2016"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Pasted from the table in document N16:
```{r}
printDetails <- FALSE
pastedData <- "TRUTH:   disease present
                       Test B is positive
                       Test B is negative
                       Test A is positive
                       100
                       17
                       Test A is negative
                       8
                       15
                       TRUTH:   disease absent
                       Test B is positive
                       Test B is negative
                       Test A is positive
                       14
                       16
                       Test A is negative
                       22
                       200"
mcdata.orig <- strsplit(split = "\n", pastedData) [[1]]
```
Note that strsplit() always returns a LIST.  Thus the [[1]].

Now we make it into a data array. (We need the package abind, because R base forgot to include the abind() function. along=3, or better rev.along=0, makes sure each component is treated as a slice.)

```{r}
mcdata.character.matrix <- matrix(mcdata.orig, ncol=3, byrow=T)
library(abind)
mcdata.character.array <- abind(mcdata.character.matrix[1:3, ], 
                               mcdata.character.matrix[4:6, ],
                               along=3)
if(printDetails) print(mcdata.character.array)
mcdata.array <- mcdata.character.array[ -1, -1,  ]
mcdata.array <-  apply(mcdata.array, MARGIN = 1:3, as.numeric) # as.numeric does not work here! 8-(
dimnames(mcdata.array) <- list(
  A=c("Apos","Aneg"),  # ROW
  B=c("Bpos","Bneg"),  # COLUMN
  TRUTH=c("present","absent") # SLICE
)	
```
### Possible data modifications go in this chunk!
```{r}
#mcdata.array[mcdata.array==16] = 36
mcdata.array["Apos", "Bneg", "absent"] = 36  # originally 16
mcdata.array["Apos", "Bneg", "present"] = 37 # originally 17
```

Now we will reshape it a "long skinny" data frame. That will be handy for some analyses.

```{r}
mcdata.df <- expand.grid(dimnames(mcdata.array))
mcdata.df$count <- c(mcdata.array)
if(printDetails) print(mcdata.df)
```

Now we can use mcdata.array and mcdata.df to test hypotheses.

Hypothesis:  A is more likely to say "positive" than B.
One approach: Just collapse over TRUTH:
```{r}
AposBneg <- mcdata.array["Apos", "Bneg", "present"] +
  mcdata.array["Apos", "Bneg", "absent"]
if(printDetails) print(AposBneg <- ### Alternative
  sum(mcdata.df [mcdata.df$A=="Apos" & 
               mcdata.df$B=="Bneg", "count"]))
AposBneg <- sum(mcdata.array["Apos", "Bneg", ])
```

Here is the other marginalized diagonal:
```{r}
  AnegBpos <- sum(mcdata.array["Aneg", "Bpos", ])
```

McNemar test, using the binomial method:
```{r}
cat("McNemar Ptest_AMorePos = ", 
         Ptest_AMorePos <- round(digits=3, 
                            pbinom(AnegBpos, AposBneg+AnegBpos, 1/2)),
      "\n")
```
Compare with mcnemar.test-- but careful, it's TWOSIDED.
The apply() call is collapsing over the 3rd dimension (TRUTH).
```{r}
test_AMoreOrLessPos <- mcnemar.test(apply(mcdata.array, 1:2, sum) )  
Ptest_AMoreOrLessPos <- test_AMoreOrLessPos$p.value
```
This is very limited, though. One test might be much better at agreeing with the truth when it's "present" but not when it's "absent".
Another approach:  control for TRUTH by combining evidence.

Hypothesis:  A is more likely to be CORRECT.
There are several approaches.
```{r}
mcdata.combined <- mcdata.array[ , , 1] + mcdata.array[ 2:1, 2:1, 2]
dimnames(mcdata.combined) <- list(c("Aright", "Awrong"), c("Bright", "Bwrong"))
mcdata.combined
Pvalue.combined = print(mcnemar.test(mcdata.combined))$p.value
```

We can also use our data frame structure:
```{r}
mcdata.df$Acorrect <- 
  (mcdata.df$A=="Apos" & mcdata.df$TRUTH=="present") |
  (mcdata.df$A=="Aneg" & mcdata.df$TRUTH=="absent") 
mcdata.df$Bcorrect <- 
  (mcdata.df$B=="Bpos" & mcdata.df$TRUTH=="present") |
  (mcdata.df$B=="Bneg" & mcdata.df$TRUTH=="absent") 
###  Collapsing over TRUTH:
AcorrectBincorrect <- mcdata.df$Acorrect & 
  ! mcdata.df$Bcorrect
nAcorrectBincorrect <- sum(mcdata.df$count[AcorrectBincorrect])
ABdisagree <- mcdata.df$Acorrect != mcdata.df$Bcorrect
nABdisagree <- sum(mcdata.df$count[ABdisagree])
cat("(collapsing) P = ",
    1 - pbinom(nAcorrectBincorrect - 1, nABdisagree, 1/2), "\n")
```
This P is ONE-sided.
Alternatively, using the array structure:
```{r}
mcdata.correctness <- mcdata.array[ , , 1] + t(mcdata.array[ , , 2])
dimnames(mcdata.correctness) <-  list(A=c("correct", "incorrect"),
                                     B=c("correct", "incorrect"))
mcdata.correctness
mcnemar.test(mcdata.correctness)  
```
Note the use of the transpose function t() to reflect the "absent" part of the array.

If there is an imbalance in marginal for TRUTH. (and there is), 
we should try controlling for TRUTH.
The worry: if most cases are "absent", then B might be
right more often just because B says "negative" more often.

             pos vs neg  ------  A vs B

                    \\            /

                   present vs absent


To define and predict "correctness", 
 adjusting for propensity to say "positive",
 we need to make a data set double the size.
 Then we will use Poisson regression, via the function glm().
```{r}
mcdata2 <- rbind(mcdata.df, mcdata.df)
mcdata2$test <- rep(c('A','B'), each=8)
(mcdata2$correct <-  ifelse(mcdata2$test=="A", 
                           mcdata2$Acorrect, mcdata2$Bcorrect))
mcdata2 <- mcdata2[c('TRUTH','count','test','correct')]

require("MASS")
(glm.out.1 <- glm(data=mcdata2,  
    correct ~ TRUTH + test, 
    weights=count, 
    family=poisson)	)
summary(glm.out.1)
```
Now we include the interaction term.
```{r}
(glm.out.2 <- glm(data=mcdata2,  correct ~ TRUTH*test, weights=count, family=poisson) )	
summary(glm.out.2)
```

## SUMMARY

The data:
```{r}
mcdata.array
```

P value for HA:"A says positive more", via binomial test:
```{r} 
Ptest_AMorePos
``` 

P value for HA:"A or B says positive more", via chi-square test:
```{r} 
Ptest_AMoreOrLessPos
```  

P value for HA:"A and B differ in correctness":
```{r}
Pvalue.combined
```
Model fit controlling for Prob(positive | test):
```{r} 
summary(glm.out.1)$coef
```
Model fit controlling for Prob(positive | test) + interaction:
```{r}
summary(glm.out.2)$coef
```
