---
title: "random-negbinomial"
author: "Roger Day"
date: "February 8, 2016"
output: html_document
---

First, we create 10000 random draws from the negative binomial (in this case, the geometric, since size = 1).
```{r}
SampleSize = 10000 
x = rnbinom(n = SampleSize, size = 1, prob = .5)
summary(x)
range(x)
hist(x)
```
Now let's take a look at a table of results.
```{r}
xtable = table(x)
xtable
```
The top row holds all the values that were observed in the sample, in sorted order. We extract them, and plot their true probabilities along the x-axis, versus their observed frequencies along the y-axis. 
```{r}
observedValues = as.numeric(names(xtable))
plot(dnbinom(observedValues, 1, .5),
	xtable/SampleSize, pch="", log="xy")
text(dnbinom(observedValues, 1, .5),
     xtable/SampleSize,
     names(xtable))
abline(a=0, b=1)
```
The agreement seems good.

Things to learn about the coding:
* Using the function $names()$ to get the values (as character strings).
* Using the function $as.numeric()$ to convert the values from strings to numbers.
* Using the function $text()$ to write text on the plot.
* Using the function $abline()$ to draw a diagonal line.


Now let's take a look at the cumulative distribution function (CDF) at the observed values:
```{r}
plot(observedValues, 
     pnbinom(observedValues, 1, .5),
     ylim=0:1)
``` 
(Note use of the $ylim$ argument to set the y-axis limits.)

Now we compare the CDF to the cumulative number frequencies of observations.
```{r}
plot(pnbinom(observedValues, 1, .5),
	cumsum(xtable/SampleSize), pch="")
text(pnbinom(observedValues, 1, .5),
     cumsum(xtable/SampleSize),
     observedValues)
abline(a=0,b=1)
```
The function $cumsum()$ calculates the cumulative sum of a vector. Note the use of pch="" to suppress the usual symbol, so we can read the text labels.

This is kind of bunched up near (1,1), so let's take the complement, then use log-log axes:
```{r}
plot(1-pnbinom(observedValues, 1, .5),
	1-cumsum(xtable/SampleSize), log="xy", pch="")
text(1-pnbinom(observedValues, 1, .5),
     1-cumsum(xtable/SampleSize),
     observedValues)
abline(a=0,b=1)
```
Looks good.  There is some random deviation for the values of very small counts, at the lower left.


The following plot illustrates the relationship between pnbinom and qnbinom.
```{r}
plot(pnbinom(Q<- 0:6, size=1, prob=0.5), Q,
       pch="Q", cex=2, col="green")

points(temp<-seq(0,1,length.out=101),
       qnbinom(temp, 1, 0.5))

title("pnbinom and qnbinom are inverse functions--\non the sample space (Q, the quantiles)")
```