---
title: "Simple Bayes decision plots"
author: "Roger Day"
date: "January 12, 2015"
output: html_document
runtime: shiny
---

 
This "shiny app" shows Bayes expected loss for decisions 

* "T": Treat
* "W": Wait

The default values for all parameters are:

Parameter  | Default value
---------  | -------------
prevalence | 0.001
sensitivity| 0.99
specificity| 0.99
Loss(Sick, Wait) |  1
Loss(Healthy, Treat |  1

The vertical blue lines occur at the default values.

* Where the W line is below the T line, the better decision is `W`.
* Where the T line is below the W line, the better decision is `T`.
* Where (if) the W and T lines cross, the two decisions are equivalent, because they have the same Bayes expected loss:

Bayes expected loss = $\rho$ = $\rho (a|X) = {E_{\theta |X}}(L(\theta ,a))$.


**To make this document work, the file "simplebayesdecision.R" must be in the same folder. **
This will make the function plot.simple.bayes() available,
 through the line  `source("simplebayesdecision.R)"` in this document. You can explore the coding in this file on your own.)

INTERACTIVE CONTROLS:

* "Parameter to vary":  the parameter along the horizontal axis.
* "Observation":  Either P for positive test, or N for negative test. Results are conditional on the observation.


## Shiny Inputs and Outputs

```{r, echo=FALSE}
source("simplebayesdecision.R")
```

```{r, echo=FALSE}
longparamnames = c("prevalence","sensitivity","specificity",
               "Loss_given_S_and_W","Loss_given_H_and_T")
paramArgDefaults = unlist(formals(plot.simple.bayes)[1:5])
paramArgNames = names(paramArgDefaults)
inputPanel(
  selectInput("parameterInputID", label = "Parameter to vary:",
              choices = longparamnames, selected = longparamnames[1]),
  selectInput("dataInputID", label="Observation",
              choices = c("P","N"), selected="P")
  )

renderPlot({
  ## First, reset all parameters to their defaults.
  sapply(paramArgNames, function(param) 
    assign(param, paramArgDefaults[param], envir=.GlobalEnv))
  ### Here pos=1 is the same as env=.GlobalEnv -- try search().
  ### Now, swap in the vector for the selected parameter.
  which.varies = which(input$parameterInputID == longparamnames)
  cat(which.varies, longparamnames[which.varies], "  ",
      paramArgNames[which.varies], "\n")
  horiz.label = input$parameterInputID 
  horiz.values = switch(paramArgNames[which.varies],
                        L.SW=seq(0.5,20,1),
                        L.HT=seq(0,1,0.05),
                        prev=1/10^seq(0.5,4,0.1),
                        sens=seq(0.1,1,.01),
                        spec=seq(0.99,1,.0002))
  #Save the default value, and reassign the vector of values
  default.value = paramArgDefaults[which.varies]
  # You need to use "assign"; we doncan't hard-code the variable name.
  assign(paramArgNames[which.varies],
         horiz.values)
  cat(horiz.label, default.value, horiz.values, "\n")
#  assign(input$parameterInputID, horiz.values)
  E.loss = simple.bayes.vectorized(
    prev, sens, spec, L.SW, L.HT, input$dataInputID)
  print(E.loss)
  E.loss.W = E.loss[ , 1]
  E.loss.T = E.loss[ , 2]
  plot(
    c(horiz.values, horiz.values), c(E.loss.W, E.loss.T), pch=" ",
    ylab="Bayes expected loss",
    xlab=input$parameterInputID,
    main=paste("Varying ", input$parameterInputID))
  lines (horiz.values, E.loss.W, pch= "W", col="red", lwd=3)
  lines (horiz.values, E.loss.T, pch= "T", col="green", lwd=3)
  points(horiz.values, E.loss.W, pch= "W", col="red", lwd=3)
  points(horiz.values, E.loss.T, pch= "T", col="green", lwd=3)
  abline(v=default.value, lty=3, lwd=5, col="blue")
  })
```


