---
title: "Simple Bayes decision plots"
output: html_document
runtime: shiny
---

 
### Using Bayes expected loss to support a medical decision. 

### Actions to choose from:

* $T$: Treat
* $W$: Wait ("Watchful Waiting")

### Bayes expected loss for action $A$ = $\rho$ = $\rho (A|X) = {E_{\theta |X}}(Loss(\theta, A))$.

* $\rho (T|X) = Loss(Healthy, Treat) \times Pr(Healthy | X)$

* $\rho (W|X) = Loss(Sick, Wait) \times Pr(Sick | X)$


### Best decisions  (Bayes rule):

* Where the $W$ line is below the $T$ line, the better decision is $W$.
* Where the $T$ line is below the $W$ line, the better decision is $T$.
* Where (if) the $W$ and $T$ lines cross, the two decisions are equivalent, because they have the same Bayes expected loss.


### You can change the default values here.


```{r, echo=FALSE}
source("simplebayesdecision.R")
```



```{r, echo=FALSE}
longparamnames = c(prev="prevalence", 
                   sens="sensitivity", spec="specificity",
               L.SW="Loss_given_S_and_W",
               L.HT="Loss_given_H_and_T")
paramArgDefaults = unlist(formals(plot.simple.bayes)[1:5])
###  a few alterations...
paramArgDefaults['prev'] = 0.1
paramArgDefaults['sens'] = 0.95
paramArgDefaults['spec'] = 0.95

paramArgNames = names(paramArgDefaults)
paramPlotRangeDefaults = c(
                        prev='1/10^seq(0.5,4,0.1)',
                        sens='seq(0.1,1,.01)',
                        spec='seq(0.99,1,.0002)',
                        L.SW='seq(0.5,20,1)',
                        L.HT='seq(0,1,0.05)')


#fluidPage(  

  inputPanel(
  br(),
  numericInput('prevalenceDefault',
               'prevalence ',
               value = paramArgDefaults['prev'],
               min = 0, max=1, step = 0.01),
  numericInput('sensitivityDefault',
               'sensitivity',
               value = paramArgDefaults['sens'],
               min = 0, max=1, step = 0.01),
  numericInput('specificityDefault',
               'specificity',
               value = paramArgDefaults['spec'],
               min = 0, max=1, step = 0.01),
  br(),
  numericInput('L.SWDefault',
               'Loss Sick&Wait ',
               value = paramArgDefaults['L.SW'],
               min = 0, max=1, step = 0.01),
  numericInput('L.HTDefault',
               'Loss Healthy&Treat ',
               value = paramArgDefaults['L.HT'],
               min = 0, max=1, step = 0.01),
  
  br(),br(),
  selectInput("parameterInputID", 
              label = "Parameter to vary:",
              choices = longparamnames, 
              selected = longparamnames[1]),
  selectInput("dataInputID", label="Observation",
              choices = c("P","N"), selected="P")
  )
#)
```

* "__Parameter to vary__":  the parameter along the horizontal axis. The vertical blue line occurs at the default value.

* "__Observation__":  Either $P$ for positive test, or $N$ for negative test. Bayes expected loss for the two actions $T$ and $W$ is conditional on the observation.



```{r echo=FALSE} 

renderPlot({
  ## First, set all parameters to their defaults.
  paramArgDefaults['prev'] = input$prevalenceDefault
  paramArgDefaults['sens'] = input$sensitivityDefault
  paramArgDefaults['spec'] = input$specificityDefault
  paramArgDefaults['L.SW'] = input$L.SWDefault
  paramArgDefaults['L.HT'] = input$L.HTDefault
  sapply(paramArgNames, function(param) 
    assign(param, paramArgDefaults[param], envir=.GlobalEnv))
  ### Here pos=1 is the same as env=.GlobalEnv -- try search().
  ### Now, swap in the vector for the selected parameter.
  which.varies = names(longparamnames)[which(input$parameterInputID == longparamnames)]
  cat(which.varies, '- ',
      longparamnames[which.varies], "--  ",
       "\n")
  horiz.label = input$parameterInputID 
  horiz.value.list = sapply(
    names(paramPlotRangeDefaults),
    function(the.rangeName)
      eval(parse(text=
                   paramPlotRangeDefaults[
                     the.rangeName]))
  )
  horiz.values = horiz.value.list[[which.varies]]
  #Save the default value, and reassign the vector of values

  default.value = paramArgDefaults[which.varies]
  # You need to use "assign"; we can't hard-code the variable name.
  cat ('paramArgNames[which.varies] ',
       paramArgNames[which.varies], '\n')
  assign(which.varies,
         horiz.values)
  cat(horiz.label, 'default VALUE=',
      default.value, '\nvalues\n', 
      horiz.values, "\n")
  #  assign(input$parameterInputID, horiz.values)
  E.loss = simple.bayes.vectorized(
    prev, sens, spec, L.SW, L.HT, input$dataInputID)
  cat('E.loss', '\n')
  print(E.loss)
  E.loss.T = E.loss[ , 1]
  E.loss.W = E.loss[ , 2]
  plot(
    c(horiz.values, horiz.values), 
    c(E.loss.T, E.loss.W), 
    pch=" ",
    ylab="Bayes expected loss",
    xlab=input$parameterInputID,
    main=paste("Varying ", input$parameterInputID))
  lines (horiz.values, E.loss.W, pch= "W", col="red", lwd=3)
  lines (horiz.values, E.loss.T, pch= "T", col="green", lwd=3)
  points(horiz.values, E.loss.W, pch= "W", col="red", cex=4)
  points(horiz.values, E.loss.T, pch= "T", col="green", cex=4)
  abline(v=default.value, lty=3, lwd=10, col="blue")
  })
  
```


