Demonstration of the three diagnostic tables.
====

### Table 1: Probabilities of test result 
This is the *model family*
```{r}
diagnosticModelFamily = rbind(c(0.95, 0.03, 0.02), 
                   c(0.03, 0.95, 0.02)) 
parameterSpace = c("healthy","sick")
sampleSpace = c("negative","positive","indeterminate")
dimnames(diagnosticModelFamily) = list(parameterSpace, sampleSpace)
print(diagnosticModelFamily)
```
The sensitivity is `r diagnosticModelFamily["sick", "positive"]`.
The specificity is `r diagnosticModelFamily["healthy", "negative"]`.

Here are the likelihood ratios:
```{r}
LRexpressions = paste0(diagnosticModelFamily[1,], '/',
                      diagnosticModelFamily[2,])
options(digits=2)
sapply (LRexpressions, function(ex) 
        eval(parse(text=ex)) )
```
### Table 2: Joint probabilities
Now we get Table 2 by multiplying each row by its probability.
```{r}
prior = c(0.9, 0.1)
diagnosticJoint = matrix(prior, nrow=2, ncol=3) *
                  diagnosticModelFamily
print(diagnosticJoint)
```

### Table 3: conditional on the observations.
```{r}
normalize = function(x) x/sum(x)
predictiveProbabilities = apply(diagnosticJoint, 2,
                                normalize)
###   "2" means columns  and "1" means rows.
```


### Simulating datasets from this joint distribution.
```{r}
sampleSize = 100
```
Each sample has `sampleSize` patients.

```{r}
simulateDiagnosticDatasets = function(ignoreMe=NULL){
  simdata = rmultinom(1, sampleSize, diagnosticJoint)
  simdata = matrix(simdata, nrow=2)
  dimnames(simdata) = dimnames(diagnosticJoint) 
  simdata
}

nReps = 1000
library(plyr)
samples = laply(1:nReps, simulateDiagnosticDatasets)
str(samples)
```
(Note the use of $laply()$ in the package $plyr$. Worth learning about.)

So *samples* is an **array** object, with 3 dimensions.
We'll fancy it up a little now.
```{r}
dimnames(samples)[[1]] = 1:nReps
names(dimnames(samples)) = c("rep number", "patient truth", "test")
str(samples)
```
What is our term for "patient truth"? For "test"?

Now let's look at the variability due just to chance.
```{r}
par(mfcol=c(3,2))
for (truth in parameterSpace) {
  for (result in sampleSpace) {
    print(paste("Truth=", truth, "  Result=", result))
    print(summary(samples[ , truth, result]))
    hist(samples[ , truth, result],
         main = paste("Truth=", truth, "  Result=", result))
  }
}
```

