---
title: "spellingSTATISTICS"
author: "Roger Day"
date: "1/22/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Suppose we pick ten letters from an infinite pile of scrabble tiles, and arrange them face down.  What's the probability that they spell STATISTICS?

  
  \[{p_S}{p_T}{p_A}{p_T}{p_I}{p_S}{p_T}{p_I}{p_C}{p_S} = p_S^3p_T^3p_A^{}p_I^2p_C^{} = \prod\limits_{j = 1}^{26} {p_j^{{m_j}}} \]
  
is the probability of getting the right letters in the right order, where $j$ is the position in the word, and $m_j$ is the count for each of the 26 letters.

Suppose that each letter has chance 1/26--- sampling WITH REPLACEMENT. All $p_j$ = 1/26.


```{r}
ourword = "STATISTICS"
wordLength = nchar(ourword)
jointProbability = print( (1/26)^wordLength )
```

This is the *joint distribution* of two events.

First, you need to select the right *number* of each letter. This is the multinomial probability:

\[\Pr (\underset{\raise0.3em\hbox{$\smash{\scriptscriptstyle\thicksim}$}}{M}  = \underset{\raise0.3em\hbox{$\smash{\scriptscriptstyle\thicksim}$}}{m} ) = \left( {\begin{array}{*{20}{c}}
  n \\ 
  {\begin{array}{*{20}{c}}
  {{m_1}}&{...}&{{m_{26}}} 
\end{array}} 
\end{array}} \right)p_1^{{m_1}}...p_{26}^{{m_{26}}} = \frac{{n!}}{{\prod\limits_{j = 1}^{26} {{m_j}!} }}\;\;\prod\limits_{j = 1}^{26} {p_j^{{m_j}}} \]

is the probability of getting the right number of each letter. In this case, the required number of each letter is given by 

```{r}
byLetter = strsplit(ourword, split="")[[1]]
letterTable = print(table(byLetter))
letterFreq = rep(0, 26)
names(letterFreq) = LETTERS
letterFreq[names(letterTable)] = letterTable
print(letterFreq)
marginalProbability = dmultinom(
  x = letterFreq, #size = 10, not necessary.
  prob = rep(1/26, 26 ) )
print(marginalProbability)

```


Next, given the right number of each letter, what is the conditional probability of putting the letters in the right order?
Each permutation is equally likely, so this is the proportion of permutations which are correct:

\[\frac{{\Pr ({\text{correct letter order}})}}{{\Pr (\underset{\raise0.3em\hbox{$\smash{\scriptscriptstyle\thicksim}$}}{M}  = \underset{\raise0.3em\hbox{$\smash{\scriptscriptstyle\thicksim}$}}{m} )}} = \frac{{\prod\limits_{j = 1}^{26} {p_j^{{m_j}}} }}{{\frac{{n!}}{{\prod\limits_{j = 1}^{26} {{m_j}!} }}\;\;\prod\limits_{j = 1}^{26} {p_j^{{m_j}}} }} = \frac{{\prod\limits_{j = 1}^{26} {{m_j}!} }}{{n!}}\]

which is: 

```{r}
nDistinctLetters = length(letterTable)
conditionalProbability = print(
  prod(factorial(letterTable)) / 
    factorial(wordLength)
)
```

This makes sense, because
  $n!$ is the total number of permutations of the $n$ slots, and the numerator $\prod\limits_{j = 1}^{26} {{m_j}!}$ is the number of permutations that match the correct slots for the word STATISTICS.
   
   
Confirming on the boundary cases:

* If the word had unique letters only, this would be only 1 permutation. 

* If the word had just one letter repeated, it would be all n! permutations.




Let's check our work:

JOINT = MARGINAL * CONDITIONAL

```{r}
marginalProbability * conditionalProbability
```

Verify:
```{r}
jointProbability
```


