---
output: html_document
---
Supplement on the Wilcoxon two-sample test: exact vs approximate, & the effect of ties
---

When the data contain ties, the exact distribution of the statistic is much more difficult and the approximations are not good. This doesn’t matter much for P values > 0.001, but can be a huge effect for very very small P values. This matters a lot, because when there are many many comparisons (high-throughput biological data), small P values are frequent, including small P values generated just by chance.

In this numerical experiment, we create a data set of many many features, and one binary label.
The effects range widely from zero to 5 standard deviations.
We round to 4 places, to see what the effect of ties might be.

```{r}
sampleSize = 120
outcomeClass = rbinom(sampleSize, 1, 1/2)
nFeatures = 300
effectSizes = rnorm(nFeatures) * 2
```
```{r warning=FALSE}
do.wilcoxes = function(Xrounded) {
  X0 <- Xrounded[outcomeClass==0]
  X1 <- Xrounded[outcomeClass==1]
  data.frame(p.approx = wilcox.test(X0, X1)$p.value,
             p.correct = wilcox.test(X0, X1, correct=T)$p.value,
             p.exact = wilcox.test(X0, X1, exact=T)$p.value,
             anyTied = length(unique(Xrounded)) < sampleSize 
                  # not all unique.
   )
}
#options.saved = options(warn=-1)
p.values.list = lapply(effectSizes, function(effSize){
  X = rnorm(sampleSize) + effSize*outcomeClass
  Xrounded <<- round(X, digits=3)
  do.wilcoxes(Xrounded)
 }
)
#options(options.saved)
```
We suppressed the warning messages, using the chunk option **{r warning=FALSE}**.

Now we combine the results.
```{r}
library(plyr)
p.values.df = ldply(p.values.list)
table(p.values.df$anyTied)

```
Note: plyr is a good library to know! In "ldply", the "l" means the input is a list, and the "d" means the output is a data frame.

```{r,fig.height=4.5,fig.width=4.5}
colors = c("red","darkgreen")[p.values.df$anyTied+1]  
cex = (1:2)[p.values.df$anyTied+1]  
pch = c("O", "T")[p.values.df$anyTied+1]  
with(p.values.df, {
    plot(p.correct, p.approx, log="xy", col=colors, pch=pch, cex=cex)
    abline(a=0,b=1)
    plot(p.exact, p.approx,log="xy", col=colors, pch=pch, cex=cex)
    abline(a=0,b=1)
  }
)
```

The first plot shows that the "correct" and "approx" values are virtually identical for a sample size of 100. Note that "correct" does not mean what it says. It means only that a "continuity correction" for the normal approximation was applied, to adjust (roughtly) for the discreteness of the Wilcoxon test statistic.

The second and third plots (green points) show that the exact P value is much smaller than the approximations, as the P value becomes quite small. This difference doesn’t matter much for P values > 0.001, but can be a huge effect for very very small P values.This matters *a lot*, because when there are many many comparisons (high-throughput biological data), small P values are frequent, including small P values generated just by chance.  



When the data contain ties, the exact distribution of the statistic is much more difficult and wilcox.text will not do it. Instead it just puts out the approximation. (There is a warning message, but as noted above we have suppressed it.) 


```{r}
plot(effectSizes, p.values.df$p.approx, pch = pch, cex=1/2, col = colors, log='y')
legend(x = 'topright', legend=c('ties', 'no ties'), c('T', 'O'), col=c('darkgreen', 'red'))
plot(effectSizes, p.values.df$p.exact, pch = pch, cex=1/2, col = colors, log='y')
legend(x = 'topright', legend=c('ties', 'no ties'), pch=c('T', 'O'), col=c('darkgreen', 'red'))

```


### Studying one data set.

To illustrate, we create a data set with a pretty good effect size. We do a no-tie version and a version with a tie, almost identical.
```{r}
effSize = 2
X = rnorm(sampleSize) + effSize*outcomeClass
X[2] = X[1] + 0.0001
do.wilcoxes(X)
t(
  sapply(6:0, function(digits) {
    c(digits=digits, do.wilcoxes(round(X, digits=digits)))
    }
    )
)
```
Now we will break the ties for the digits=3 dataset. That should give a more appropriately small P value.
```{r}
Xrounded3 = round(X, digits=3)
do.wilcoxes(Xrounded3)
which(duplicated(Xrounded3))
ties = which(duplicated(Xrounded3))
Xbrokentie = Xrounded3
Xbrokentie[ties] = Xbrokentie[ties] + 1e-7
do.wilcoxes(Xbrokentie)
```
So we have regained the power lost when the ability to do the exact permutation was lost.

The 
computing time increases dramatically due to *exact=TRUE*, when there are no ties. If there are ties, the option is ignored.
```{r}
unix.time(for(i in 1:20)do.wilcoxes(Xrounded3))
unix.time(for(i in 1:20)do.wilcoxes(Xbrokentie))
```
