---
title: "Wilcoxon two-sample test: ties make a mess of things...what to do?"
author: "Roger Day"
date: "2024-08-17"
output: html_document
---

```{r setup, include=FALSE}
# https://medium.com/@rogerday_17101/wilcoxon-two-sample-test-ties-make-a-mess-of-things-what-to-do-ab9c2c7e36e6?source=friends_link&sk=deb92924331cad6e6c6411b4d35e588a

#knitr::opts_chunk$set(echo = FALSE, results='hide')
knitr::opts_chunk$set(echo = FALSE, comment = '')
```
---
Wilcoxon two-sample test: exact vs approximate, & the effect of ties
---


```{r include=FALSE,echo=FALSE}
sampleSize = 100
nFeatures = 300
DIGITS = 3
effectSD = 2
effectSizes = rnorm(nFeatures) * effectSD

```

Trigger warning: this article uncritically presumes the use of "P values" in comparing two groups of numbers. 

There are loads of reasons to despise P values; elsewhere I have a polemic chanting a few of those reasons.  
It also presumes that `multiple testing adjustments` are in play.  That's got some ridiculous paradoxes too. And a corresponding   polemic. All in good time on the [`StatisticsBeautiful`](https://www.youtube.com/@statisticsbeautiful1098) channel.

So... let's just agree to get along for the next few minutes. 


An old old work horse of "classical" frequentist statistical analysis, the Wilcoxon test comes in two forms.  This will be about the two-sample form, comparing two collections of numbers, to see if one tends to be bigger than the other (essentially, are the medians different?). Those numbers constitute a "feature".  The Wilcoxon test gets employed to judge whether the feature tends to be higher in one sample, one group, than the other.

What is a "feature"?  A number. Statisticians used to call them "covariates".  Not classy. Better PR in computer science. 

The testing tool is the function `wilcox()`, a basic workhorse in the statistical language `R`. Why was it necessary to cheat poor Mr. Wilcoxon out of the last two letters of his name?  DM me if you know! 

#### The problem with ties

The basis for the test is to count the number of pairs, one from group #1 and the other from group #2, for which the group #2 number is larger. 

When there are many many features to evaluate  (such as high-throughput genomic data), P values come spitting out in great numbers, and many very small P values can be generated just by chance. So one becomes skeptical of "discoveries", and requires extra strong evidence, an exceptionally tiny P value. That P value can be computed with a approximations, or exactly based on the randomization distribution. The exact calculation has more prestige. (Though, like many frequentist calculations, the sampling distribution assumed is not nature's sampling distribution, so "exact" is  not exactly true.)

But when the data contain ties, the exact distribution of the statistic is much more difficult and the approximations are not good. This doesn’t matter much for P values > 0.001, but can be a huge effect for very very small P values. Very very small P values matter <i>a lot</i> when the number of features, the number of comparisons, is huge.

### Numerical experiment
In this numerical experiment, I create a data set with `r sampleSize` observations and `r nFeatures` features, and one binary label per observation; no imagination here, so the labels are #1 and #2. 

For each feature, the effect size is chosen randomly, from a normal distribution with a standard deviation of `r effectSD`. That effect is added to the group #2 observations.

Let's look at three ways that `wilcox` can calculate the P value: normal approximation (`default`), `correct=TRUE`, and `exact=TRUE`.

```{r}
outcomeClass = rbinom(sampleSize, 1, 1/2) + 1
```
```{r warning=FALSE}
do.wilcoxes = function(Xrounded) {
  X1 <- Xrounded[outcomeClass==1]
  X2 <- Xrounded[outcomeClass==2]
  data.frame(p.approx = wilcox.test(X1, X2)$p.value,
             p.correct = wilcox.test(X1, X2, correct=T)$p.value,
             p.exact = wilcox.test(X1, X2, exact=T)$p.value,
             anyTied = length(unique(Xrounded)) < sampleSize 
                  ## not all unique.
   )
}
#options.saved = options(warn=-1)
p.values.list = lapply(effectSizes, function(effSize){
  X = rnorm(sampleSize) + effSize*outcomeClass
  Xrounded <<- round(X, digits=DIGITS)
  do.wilcoxes(Xrounded)
 }
)
#options(options.saved)
library(plyr)
## Note: plyr is a good library to know! In "ldply", the "l" means the input is a list, and the "d" means the output is a data frame.
p.values.df = ldply(p.values.list)
anyTiedTable = table(p.values.df$anyTied)

```
(Want to see the R markdown which created this document? Look for the file wilcoxon-ties-experiment.Rmd at https://github.com/professorbeautiful/bioinf2118. In R markdown,  the warning messages have been suppressed using the chunk option **{r warning=FALSE}**. Warnings are so bourgeois.)

Now we combine the results across all the `r nFeatures` features. Here the rounding takes the `r nFeatures` features and creates ties in `r anyTiedTable[2]` of them, with `r anyTiedTable[1]` still without ties.


#### Figure 1
```{r figure1,fig.height=4.5,fig.width=4.5, cache=FALSE}
colors = c("red","darkgreen")[p.values.df$anyTied+1]  
cex = c(1,1)[p.values.df$anyTied+1]  
pch = c("O", "T")[p.values.df$anyTied+1]  
with(p.values.df, {
    plot(p.approx, p.correct, log="xy", col=colors, pch=pch, cex=cex, 
         xlab="P value: normal approximation", 
         ylab="P value: \"corrected\" approximation")
    abline(a=0,b=1)
    title('"correct=TRUE" (continuity correction)')
  legend(x = 'topleft', legend=c('feature with ties', 'feature with NO ties'), pch=c('T', 'O'), col=c('darkgreen', 'red'))
}

)
```


In Figure 1, each point is a feature, with the effect size chosen randomly.
The "green T" is a feature that has at least one tie.
The "red O" is a feature that has no ties.
The graph compares to approximation values to the correct values. Excuse me, the `correct=TRUE` P values. 
"Correct" does not mean they are correct. 
It means only that a "continuity correction" for the normal approximation was applied, to adjust (roughly) for the discreteness of the Wilcoxon test statistic.  Sounds good; but makes no discernable difference, whether there are ties in the feature or not.

### The Exact-o knife


Let's instead compare the approximation to the exact P value. Excuse me, the `exact=TRUE` P value. Surprise, it may not be exact, as we'll see.



#### Figure 2

```{r figure2, cache=FALSE}
with(p.values.df, {
    plot(p.approx,p.exact, log="xy", col=colors, pch=pch, cex=cex,
         xlab="P value: normal approximation", 
         ylab="P value: \"exact\" (sometimes)")
    abline(a=0,b=1)
    title('"exact=TRUE" ')
    legend(x = 'topleft', legend=c('feature with ties', 
                                   'feature with NO ties'),
           pch=c('T', 'O'), col=c('darkgreen', 'red'))

  }
)
```
Now this is strange!   For features with ties, the green T's agree perfectly like before.  But for those with no ties, when the exact P value is really small, the approximation is terrible, hugely bigger, less significant, than it should be.

This matters *a lot*, because when there are many many comparisons (high-throughput biological data), small P values are frequent, including small P values generated just by chance. People apply very stringent P value cutoffs to reduce the risk of false positive "discoveries" that are just artifacts of chance.

What's going on??

## The effect of ties

When the data contain ties, the exact distribution of the statistic is much MUCH more computationally intensive, and `wilcox.test` simply will not do it. Instead it just puts out the approximation. (There is a warning message, but as noted above we have suppressed it.) 


Let's look at this by effect size.
For the approximate P values, ties and no-ties features follow the same trajectory.

#### Figure 3

```{r figure3}
plot(effectSizes, p.values.df$p.approx, pch = pch, cex=1/2, col = colors, log='y')
legend(x = 'topright', legend=c('ties', 'no ties'), pch=c('T', 'O'), col=c('darkgreen', 'red'))
```

But for the "exact" computation, the presence of ties prevents the "exact" method from being used.

#### Figure 4


```{r figure4}
plot(effectSizes, p.values.df$p.exact, pch = pch, cex=1/2, col = colors, log='y')
legend(x = 'topright', legend=c('ties', 'no ties'), pch=c('T', 'O'), col=c('darkgreen', 'red'))

```


#### Studying one feature.

```{r}
effSize = 2

```

To illustrate, we create a single feature with a pretty strong effect size =  `r effSize`. We do a no-tie version, but where the second value is replaced by X[2] = X[1] + 0.0001, very close to a tie.


```{r}
X = rnorm(sampleSize) + effSize*outcomeClass
X[2] = X[1] + 0.0001   #almost a tie
wilcox.out = do.wilcoxes(X)
```

Then we successively round the data to fewer and fewer digits.

#####  Changing the rounding
```{r warning=FALSE, results='show'}
print.data.frame(
  row.names=FALSE,
  as.data.frame(t(
  sapply(6:0, function(digits) {
    c(digits=digits, do.wilcoxes(round(X, digits=digits)))
    }
    )
)))
```
Look at the "p.exact" column.
Once you round to 3 or fewer digits, ties appear. And all hell breaks loose.  In the sense that the "exact" calculation is no longer done, but the approximation substitutes for it. 

Is that good?  Well, if you don't mind increasing the P value by  orders of magnitude...   But if you are committed to coping with multiple testing with some "significance" adjustment like the Bonferroni method, then that is a disaster.

Elsewhere, I have a long rap about the perils of the P value, and another long rap about the perils of multiple testing, so ... calm down.


### Breaking the ties

Now we will take the digits=3 rounded feature, but break the ties, by adding a tiny increment to a duplicated value. That should give a more appropriately small P value.

```{r warning=TRUE, results='hide'}
Xrounded3 = round(X, digits=3)
wilc.Xrounded3 = do.wilcoxes(Xrounded3)
which(duplicated(Xrounded3))
ties = which(duplicated(Xrounded3))
Xbrokentie = Xrounded3
Xbrokentie[ties] = Xbrokentie[ties] + 1e-7
wilc.Xbrokentie = do.wilcoxes(Xbrokentie)
```

Oops, I forgot to suppress the warning message this time. But...
Aha!  with the original data with ties, the P value is `r signif(digits=2, wilc.Xrounded3$p.exact)`, but with the teensy breaking of the ties, it's `r signif(digits=2, wilc.Xbrokentie$p.exact)`.  Now the exact P value is orders of magnitude smaller, as it should be.
So we have regained the power lost when the ability to do the exact permutation was lost.

The documentation for `wilcox` says:
<pre>
For stability reasons, it may be advisable 
to use rounded data or to set digits.rank = 7, say, ...
</pre>
That might not be such good advice.

### Computing time
The computing time increases somewhat due to *exact=TRUE*, when there are no ties. *If there are ties, the option is ignored*. We suppress the warnings again.

```{r warning=FALSE}
nREPS = 200
timingFor.Xrounded3 = system.time(for(i in 1:nREPS)do.wilcoxes(Xrounded3))['user.self']
timingFor.Xbrokentie = system.time(for(i in 1:nREPS)do.wilcoxes(Xbrokentie))['user.self']
```

For looping over `r nREPS` repetitions,
the computation takes `r timingFor.Xrounded3*1000` milliseconds for the "rounded to 3 digits" feature, with its ties, and  `r timingFor.Xbrokentie*1000` milliseconds when the ties are broken by tiny arbitrary amounts.

There is an alternative to the built-in `wilcox.test` function. In the R package 
[`coin`](https://search.r-project.org/CRAN/refmans/coin/html/00Index.html), the function `wilcox_test` is worth a look (notice the underscore instead of the dot; cool, ... but still the `"on"` is missing!? No matter; cyberspace shows no WilcoxOFF). 
